---
title: "Reproducible Research: Assignment 1"
author: "Daniel Thornton"
date: "10/07/2018"
output:
  html_document: default
  pdf_document: default
---

This document processes and describes data taken from personal movement using activity monitoring devices

## Setup

Set default options and install the right libraries

```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
library(lattice) 

```

## Load data

Read in the data and include the necessary libraries, assuming the activity.zip exists in the current working directory

```{r}
library(lattice) 

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile = paste0(getwd(), '/activity.zip'), method = "curl")
unzip("activity.zip",exdir = "data")
data <- read.csv("activity.csv")

```

## Daily activity

```{r}
# Histogram of the steps per day
steps_per_day <- aggregate(data[, c('steps')], list(data$date), FUN = function(x) sum(x, na.rm = TRUE))
colnames(steps_per_day) <- c('Date','Steps')
hist(steps_per_day$Steps, xlab="steps per day", breaks=40, main="Histogram of total steps per day")
```

```{r}
mean_steps_per_day <- mean(steps_per_day$Steps)
median_steps_per_day <- median(steps_per_day$Steps)
```

Mean steps in the data per day: `r mean_steps_per_day`

Median steps in the data per day: `r median_steps_per_day`

## Interval activity

```{r}
# Time series of steps per 5 minute window, averaged across all days
steps_per_window <- aggregate(data[, c('steps')], list(data$interval), FUN = function(x) mean(x, na.rm = TRUE))
colnames(steps_per_window) <- c('interval','mean_steps')

interval_with_most_steps <- steps_per_window[steps_per_window$mean_steps == max(steps_per_window$mean_steps),]$interval

plot(steps_per_window, type="l", xlab="5 min time interval starting from 00:00 hours", ylab="Steps")
```

The interval with the largest mean number of steps is interval no. `r interval_with_most_steps`

## Finding and imputating missing values

```{r}
# How many rows with NAs in the original data
rows_with_missing_data <- nrow(data[!complete.cases(data), ])
```

There are `r rows_with_missing_data` rows with missing values for the step count.

To patch the missing values, the average for that time interval taken from the other days is used, recorded in the steps_per_window variable, and joined onto the original data. 

```{r}
# Patch up the rows with missing data and plot a histogram of the new data

rows_with_missing_data <- data[!complete.cases(data$steps), ]
rows_without_missing_data <- data[complete.cases(data$steps), ]

rows_with_missing_data <- merge(rows_with_missing_data, steps_per_window, by.x="interval", by.y="interval")
rows_with_missing_data$steps <- as.integer(round(rows_with_missing_data$mean_steps))
rows_with_missing_data <- rows_with_missing_data[,c(1:3)]

new_data <- rbind(rows_with_missing_data, rows_without_missing_data)
new_data <- new_data[order(new_data$date,new_data$interval),]

```

```{r}
# Plot steps per day using the new imputed data
steps_per_day_new <- aggregate(new_data[, c('steps')], list(new_data$date), FUN = function(x) sum(x, na.rm = TRUE))
colnames(steps_per_day_new) <- c('date','steps')
hist(steps_per_day_new$steps, xlab="steps per day", breaks=40, main="Histogram of total steps per day with imputation")
```

```{r}
# Check the mean and median of the new imputed data
new_total_steps <- sum(steps_per_day_new$steps, na.rm=TRUE)
new_mean_steps <- mean(steps_per_day_new$steps, na.rm=TRUE)
new_median_steps <- median(steps_per_day_new$steps, na.rm=TRUE)

```

After imputing missing values where the data on step count was missing:

The mean number of steps is: `r new_mean_steps`

The median number of steps is: `r new_median_steps`

## Compare weekday and weekend activity

Plot the comparison of weekend activity to weekday activity, based on 5 minute windows of activity recording.

```{r warning=FALSE}
# Use a factor for the day of week

new_data$day <- weekdays(as.Date(new_data$date))
weekends <- new_data[new_data$day %in% c('Saturday','Sunday'),]
weekends$day <- factor('weekend')
weekdays <- new_data[new_data$day %in% c('Monday','Tuesday','Wednesday','Thursday','Friday'),]
weekdays$day <- factor('weekday')
new_data_new <- rbind(weekends,weekdays)

weekdays_steps_per_day <- aggregate(weekdays[, c('steps')], list(weekdays$interval), FUN = function(x) mean(x, na.rm = TRUE))
colnames(weekdays_steps_per_day) <- c('interval','mean_steps')
weekdays_steps_per_day$day_type <- c('weekday')

weekend_steps_per_day <- aggregate(weekends[, c('steps')], list(weekends$interval), FUN = function(x) mean(x, na.rm = TRUE))
colnames(weekend_steps_per_day) <- c('interval','mean_steps')
weekend_steps_per_day$day_type <- c('weekend')

by_day_type <- rbind(weekdays_steps_per_day, weekend_steps_per_day)

xyplot(by_day_type$mean_steps~by_day_type$interval | by_day_type$day_type, xlab="Interval", ylab="Number of steps",layout=c(1,2), type="l", title="Plot of the number of steps per 5 minute window comparing weekend to weekdays.")

```

